---
title: "Exploring Gun Deaths in America"
author: "Abigail Bergman"
output: github_document
---

## Get the data

```{r}
# don't modify this chunk unless you still need to install rcfss
# if so, run "devtools::install_github("uc-cfss/rcfss")" in the console first

library(tidyverse)    # load tidyverse packages, including ggplot2 
library(rcfss)

# load the data
data("gun_deaths")
gun_deaths
```

## Generate a data frame that summarizes the number of gun deaths per month.
library(tidyverse)
library(knitr)
library(rcfss)

count(gun_deaths, month)

### Print the data frame as a formatted `kable()` table.

```{r}
(month_count <- count(gun_deaths, month) %>%
kable(caption = "Gun Deaths in the US from 2012-14, shown by month", col.names = c("Month","Number of Gun Deaths")))
```

### Generate a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r}
#Rename months 
month <- c(1,2,3,4,5,6,7,8,9,10,11,12)
month.abb[month]

ggplot(gun_deaths, aes(month), ylab = "Number of Gun Deaths") + geom_bar() 

#Below chunk isnt working  
  scale_x_discrete(name="month", breaks = c("1","2","3","4","5","6","7","8","9","10","11","12"), labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) 
  
  
(by_month <- gun_deaths %>%
  drop_na(month) %>%
  ggplot(aes(x=month)) +geom_bar()) + labs(title="Gun deaths by month")

```

## Generate a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r}


gun_deaths %>%
   arrange(desc(intent)) %>%
  drop_na(intent) %>%
  ggplot(aes(x=intent)) +geom_bar() + labs(title = "Number of gun deaths by intent")


#Still trying to reorder graph
```

## Generate a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r}
#make boxplot
gun_deaths %>%
  ggplot(aes(x=sex, y =age))+geom_boxplot()

#print female average

group_by(gun_deaths, sex)%>%
  filter(sex == "F") %>%
  summarize(avg_age = mean(age, na.rm=TRUE)) 

```

## How many white males with at least a high school education were killed by guns in 2012?

```{r}
#COME BACK TO THIS
filter(gun_deaths, education != "Less than HS") %>%
  group_by(race == "white") %>%
  summarize(white_males = sex == "M", na.rm=TRUE)


```

## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r}
# convert continuous to categorical
(season <- gun_deaths%>% 
  mutate(Season=cut(month, breaks=c("0", "3", "6", "9", "12"), labels=c("Winter","Spring","Summer", "Fall")))) 

group_by(gun_deaths, Season) %>% 
  count
#STILL NEED TO FIND SEASON WITH MOST
 

```

# Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r}

```

## Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r}

```

## Session info

```{r}
# don't modify this chunk
devtools::session_info()
```


