---
title: "Exploring Gun Deaths in America"
author: "Abigail Bergman"
output: github_document
---

## Get the data

```{r}
# don't modify this chunk unless you still need to install rcfss
# if so, run "devtools::install_github("uc-cfss/rcfss")" in the console first

library(tidyverse)    # load tidyverse packages, including ggplot2 
library(rcfss)

# load the data
data("gun_deaths")
gun_deaths
```

## Generate a data frame that summarizes the number of gun deaths per month.
```{r}
library(tidyverse)
library(knitr)
library(rcfss)

count(gun_deaths, month) 

```

### Print the data frame as a formatted `kable()` table.

```{r}
(month_count <- count(gun_deaths, month) %>%
kable(caption = "Gun Deaths in the US from 2012-14, shown by month", col.names = c("Month","Number of Gun Deaths")))
```

### Generate a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r}
#Rename months 

(by_month <- gun_deaths %>%
  drop_na(month) %>%
  ggplot(aes(x=month)) +geom_bar()) + labs(title="Gun Deaths by Month, 2012-14", x = "Month", y = "Gun Deaths") + scale_x_discrete(limits = c("1","2","3","4","5","6","7","8","9","10","11","12"), labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
  

```

## Generate a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r}
#make new data frame reordering 
(n_intent <- gun_deaths %>%
  drop_na(intent) %>%
  count(intent))

#make column chart
  ggplot(n_intent, aes(x=reorder(intent, -n), y = n)) + geom_col() + labs(title = "Number of gun deaths by intent 2012-14", x = "Intent", y = "Number of deaths") 


```

## Generate a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r}
#make boxplot
gun_deaths %>%
  ggplot(aes(x=sex, y =age)) +
  geom_boxplot() + labs(title = "Boxplot of Age of Gun Death Victims by Sex 2012-14")

#print female average

group_by(gun_deaths, sex)%>%
  filter(sex == "F") %>%
  summarize(avg_age = mean(age, na.rm=TRUE))

print("Female Average = 43.69507")

```

## How many white males with at least a high school education were killed by guns in 2012?

```{r}
filter(gun_deaths, year== 2012, sex =="M", race == "White") %>%
  group_by(education) %>%
  count()

total <- (7912+4258+3029)

print("15199 whites males with at least a high school education were killed by guns in 2012")


```

## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r}
# convert continuous to categorical
(season <- gun_deaths%>% 
  mutate(Season=cut(month, breaks=c("0", "3", "6", "9", "12"), labels=c("Winter","Spring","Summer", "Fall"))) %>%
  group_by(Season) %>%
   count())

ggplot(season, aes(x=Season, y=n)) + geom_col() + labs(title = "Gun Deaths by Season, 2012-14", y = "Gun deaths") 

print("Summer had the most gun deaths with 26280")


```

# Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r}

#Better graph
(likely_all <- gun_deaths %>%
  group_by(race, intent) %>%
  count())

ggplot(likely_all, aes(x = race, y = n, fill = intent)) + geom_col(position = "fill") + labs(title = "Comparison of Homicide vs. Suicide over Whites, Blacks, and Hispanics", x = "Race", y = "Gun Deaths") + scale_y_continuous(labels = scales::percent)

print("Whites are more likely to die from suicide, Blacks are more likely to die from homicide, Hispanics are more likely to die from homicide")


```

## Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r}
#make a table showing number of gun deaths associated with police and not associated with police

gun <- gun_deaths %>%
  mutate(bypolice = ifelse(police %in% 0, "Non Police Involved","Police Involved"))
 
gun %>%
  group_by(bypolice)%>%
  count()

print("There are 99396 non police involved gun deaths and 1402 police involved gun deaths")

#gun deaths and age
ggplot(gun, aes(x=bypolice, y = age)) + geom_boxplot() + labs(title = "Police Involved vs Non Police Involved Deaths, By Age, 2012-14", y = "Age of Victim", x = "Shooting Type")


#comparison by race

(race_police <- gun %>%
  group_by(race, bypolice) %>%
  count())

ggplot(race_police, aes(x = bypolice, y = n, fill = race)) + geom_col(position = "fill") + labs(title = "Police Involved vs Non Police Involved Deaths, By Race, 2012-14", y = "Percentage of Gun Deaths", x = "Shooting Type") + scale_y_continuous(labels = scales::percent)
                                                                      #All three variables

ggplot(gun, aes(x=race, y = age)) + geom_boxplot(aes( fill = bypolice)) + coord_flip() + guides(fill = guide_legend(title = "Shooting type")) + labs(title = "Race and Age by Shooting Type", x = "Age", y = "Race")

ggplot(gun, aes(x=bypolice, y = age)) + geom_boxplot(aes( fill = race)) + guides(fill = guide_legend(title = "Race")) + labs(x = "Shooting Type", y = "Age", title = "Shooting Type by Age and Race")

```

## Session info

```{r}
# don't modify this chunk
devtools::session_info()
```


